---
title: "Gaussian Processes"
author: "Matthew Edwards"
date: "28/09/2021"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(mvtnorm)
```

## Content

The Gaussian Process is a natural extension of the multivariate Gaussian distribution which is a natural extension of the Gaussian distribution. As such the Gaussian process is going to be introduced in a series of steps:

1. Gaussian distribution
2. Bivariate Gaussian distribution
3. Multivariate Gaussian distribution
4. Gaussian Process distribution

## One Variable Data

In the following table $X$ is a variable. Each row corresponds to a sample from that variable.

```{r}
data = tibble(X = rnorm(5))
data
```

The **sample mean** and the **sample variance** are important properties of these data.

## Gaussian Distribution

The Gaussian distribution describes the probabilistic properties of a variable $X$. The Gaussian distribution is parameterised by a mean and variance:

- The **mean** $\mu$ is the centre of $X$ values
- The **variance** $\sigma^2$ is the spread of $X$ values

If we assume a variable $X$ has a Gaussian distribution with mean $\mu$ and variance $\sigma^2$ then $X\sim N(\mu,\sigma^2)$.

## Gaussian Distribution

- $\mu=0$
- $\sigma^2=1$

```{r}
ggplot(tibble(x = c(-3, 3)), aes(x)) + 
  stat_function(fun = dnorm) +
  labs(x = "X", y = "Probability Density")
```

## Gaussian Distribution

- $\mu=-1$ (decrease mean)
- $\sigma^2=1$

```{r}
ggplot(tibble(x = c(-3, 3)), aes(x)) + 
  stat_function(fun = dnorm, col = "grey") +
  stat_function(fun = dnorm, args = list(mean = -1)) +
  labs(x = "X", y = "Probability Density")
```

## Gaussian Distribution

- $\mu=0$
- $\sigma^2=1$

```{r}
ggplot(tibble(x = c(-3, 3)), aes(x)) + 
  stat_function(fun = dnorm) +
  labs(x = "X", y = "Probability Density")
```

## Gaussian Distribution

- $\mu=0$
- $\sigma^2=2$ (increase variance)

```{r}
ggplot(tibble(x = c(-3, 3)), aes(x)) + 
  stat_function(fun = dnorm, col = "grey") +
  stat_function(fun = dnorm, args = list(sd = sqrt(2))) +
  labs(x = "X", y = "Probability Density")
```

## Two Variable Data

In the following table $X_1$ and $X_2$ are variables. Each row corresponds to a sample from these variables.

```{r}
data = tibble(X1 = rnorm(5), X2 = rnorm(5))
data
```

The **sample correlation** is an important property of these data.

## Bivariate Gaussian Distribution

The bivariate Gaussian distribution describes the probabilistic properties of two variables $X_1$ and $X_2$. The bivariate Gaussian distribution is parameterised by a mean and variance for each variable and a correlation for the pair of variables:

- The **mean** $\mu_t$ is the centre of $X_t$ values
- The **variance** $\sigma^2_t$ is the spread of $X_t$ values
- The **correlation** $\rho$ is the association of $X_1$ and $X_2$ values

## Bivariate Gaussian Distribution

The means $\mu_1$ and $\mu_2$ can be collected into a **mean vector**:

$$
\boldsymbol\mu=
\begin{bmatrix}
\mu_1\\
\mu_2
\end{bmatrix}
$$

The variances $\sigma^2_1$ and $\sigma^2_2$ and correlation $\rho$ can be collected into a **variance matrix**:

$$
\boldsymbol\Sigma=
\begin{bmatrix}
\sigma_1^2 & \rho\sigma_1\sigma_2\\
\rho\sigma_2\sigma_1 & \sigma_2^2
\end{bmatrix}
$$

If we assume the variables $X_1$ and $X_2$ have a bivariate Gaussian distribution with mean vector $\boldsymbol\mu$ and variance matrix $\boldsymbol\Sigma$ then $(X_1,X_2)\sim N_2(\boldsymbol\mu,\Sigma)$.

## Bivariate Gaussian Distribution

$$
\boldsymbol\Sigma=
\begin{bmatrix}
1 & 0\\
0 & 1
\end{bmatrix}\quad\text{(correlation of zero)}
$$

```{r warning=FALSE}
ggplot(tibble(x = rnorm(1000), y = rnorm(1000)), aes(x, y)) +
 geom_point() + 
  geom_density_2d_filled(alpha = 0.5) +
  xlim(-4, 4) +
  ylim(-4, 4) +
  labs(x = "X1", y = "X2")
```

## Bivariate Gaussian Distribution

$$
\boldsymbol\Sigma=
\begin{bmatrix}
1 & 0\\
0 & 0.5
\end{bmatrix}\quad\text{(decrease X2 variance)}
$$

```{r warning=FALSE}
ggplot(tibble(x = rnorm(1000), y = rnorm(1000, sd = sqrt(0.5))), aes(x, y)) +
 geom_point() + 
  geom_density_2d_filled(alpha = 0.5) +
  xlim(-4, 4) +
  ylim(-4, 4) +
  labs(x = "X1", y = "X2")
```

## Bivariate Gaussian Distribution

$$
\boldsymbol\Sigma=
\begin{bmatrix}
1 & 0.8\\
0.8 & 1
\end{bmatrix}\quad\text{(increase correlation)}
$$

```{r warning=FALSE}
mat = rmvnorm(1000, sigma = matrix(c(1, 0.8, 0.8, 1), nrow = 2)) 
colnames(mat) = c("x", "y")
as_tibble(mat) %>% 
  ggplot(aes(x, y)) +
  geom_point() + 
  geom_density_2d_filled(alpha = 0.5) +
  xlim(-4, 4) +
  ylim(-4, 4) +
  labs(x = "X1", y = "X2")
```

## Multivariate Gaussian Distribution

The multivariate Gaussian distribution describes the probabilistic properties of $d$ variables $X_1,\dots,X_d$. The multivariate Gaussian distribution is parameterised by a mean and variance for each variable and a correlation for each pair of variables:

- The **mean** $\mu_t$ is the centre of $X_t$ values
- The **variance** $\sigma^2_t$ is the spread of $X_t$ values
- The **correlation** $\rho_{st}$ is the association of $X_s$ and $X_t$ values

## Multivariate Gaussian Distribution

The means $\mu_1,\dots,\mu_d$ can be collected into a  **mean vector**:

$$
\boldsymbol\mu=
\begin{bmatrix}
\mu_1\\
\vdots\\
\mu_d
\end{bmatrix}
$$

The variances $\sigma^2_1,\dots,\sigma^2_d$ and correlations $\rho_{st}$ can be collected into a **variance matrix**:

$$
\boldsymbol\Sigma_{st}=\rho_{st}\sigma_s\sigma_t
$$

If we assume the variables $X_1,\dots,X_d$ have a multivariate Gaussian distribution with mean vector $\boldsymbol\mu$ and variance matrix $\boldsymbol\Sigma$ then $(X_1,\dots,X_d)\sim N_d(\boldsymbol\mu,\boldsymbol\Sigma)$.

## Alternative Plot

Plotting samples from a multivariate ($d>2$) Gaussian distribution is impossible. This would require the screen to be three dimensional!

However, they can be plotted with an **index plot**.

## Index Plot ($d=2$)

$\rho_{st}=\exp(|s-t|^2/200)\quad\text{(unit variance)}$

```{r}
make_sigma = function(d, l=1) {
 sigma = matrix(rep(0, d * d), nrow = d)
 for(i in seq_len(d)) {
   for(j in seq_len(d)) {
     sigma[i, j] = exp(-abs(i-j)^2/(2*l^2))
   }
 }
 return(sigma)
}
```

```{r warning=FALSE}
n = 10
d = 2
mat = rmvnorm(n, sigma = make_sigma(d, 10)) 
colnames(mat) = as.character(seq_len(d))
as_tibble(mat) %>% 
  mutate(sample = factor(seq_len(n))) %>% 
  pivot_longer(cols = as.character(seq_len(d)), names_to = "index") %>% 
  mutate(index = as.integer(index)) %>% 
  ggplot(aes(index, value, col = sample)) +
  geom_point() +
  geom_line() +
  labs(x = "Index", y = "Value", col = "Sample") +
  xlim(1, 5) +
  ylim(-3, 3)
```

## Index Plot ($d=5$)

$\rho_{st}=\exp(|s-t|^2/200)\quad\text{(unit variance)}$

```{r warning=FALSE}
n = 10
d = 5
mat = rmvnorm(n, sigma = make_sigma(d, 10)) 
colnames(mat) = as.character(seq_len(d))
as_tibble(mat) %>% 
  mutate(sample = factor(seq_len(n))) %>% 
  pivot_longer(cols = as.character(seq_len(d)), names_to = "index") %>% 
  mutate(index = as.integer(index)) %>% 
  ggplot(aes(index, value, col = sample)) +
  geom_point() +
  geom_line() +
  labs(x = "Index", y = "Value", col = "Sample") +
  xlim(1, 5) +
  ylim(-3, 3)
```

## Index Plot ($d=50$)

$\rho_{st}=\exp(|s-t|^2/200)\quad\text{(unit variance)}$

```{r warning=FALSE}
n = 10
d = 50
mat = rmvnorm(n, sigma = make_sigma(d, 10)) 
colnames(mat) = as.character(seq_len(d))
as_tibble(mat) %>% 
  mutate(sample = factor(seq_len(n))) %>% 
  pivot_longer(cols = as.character(seq_len(d)), names_to = "index") %>% 
  mutate(index = as.integer(index)) %>% 
  ggplot(aes(index, value, col = sample)) +
  geom_point() +
  geom_line() +
  labs(x = "Index", y = "Value", col = "Sample") +
  xlim(1, 50) +
  ylim(-3, 3)
```

## Gaussian Process

Notice that samples from the $50$-dimensional multivariate Gaussian distribution are discrete functions over the index $t$ from 1 to $d$. 

Is it possible to sample continuous functions over an index $t\in\mathbb{R}$. This requires a **Gaussian Process** distribution. 

## Gaussian Process

1. Multivariate Gaussian distribution requires a:
    + mean $\mu_t$ for each $t$ from 1 to $d$. 
    + variance $\sigma^2_t$ for each $t$ from 1 to $d$
    + correlation $\rho_{st}$ for each $s,t$ from 1 to $d$
2. Gaussian Process distribution requires a:
    + mean $\mu(t)$ for each $t\in\mathbb{R}$
    + variance $\sigma^2(t)$ for each $t\in\mathbb{R}$
    + correlation $\rho(s,t)$ for each $s,t\in\mathbb{R}$
  
## Gaussian Process  

- $\mu(t)$ is the mean function
- $\sigma^2(t)$ is the variance function
- $\rho(s,t)$ is the correlation function

If we assume a function $X(t)$ has Gaussian Process distribution with mean function $\mu(t)$, variance function $\sigma^2(t)$ and correlation function $\rho(s,t)$ then $X(t)\sim N_d(\mu(t),\sigma(s)\sigma(t)\rho(s,t))$.

[Gaussian Process Playground](https://durrande.shinyapps.io/gp_playground/)

*In the index plots we used a squared exponential covariance function with $\sigma^2(t)=1$ and $\theta=0.02$.*